When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-02-14 doc/paper/vpd_et_paper.tex
done with pierre's comments, still ahve some q's (mostly with citations), and then just clean up, format to Glob Change Bio, and submit

$\Delta$'s presence in the scaling term also matches physical intuition. Evaporative cooling will dampen the ability of the atmosphere to take more moisture (UNCLEAR SENTENCE REFORMULATE), because $e_{s}$ decreases with decreasing temperature. The decrease in $e_{s}$ is proportional to $\Delta$ ($\delta e_{s} = \Delta \delta T$). So as $\Delta$ increases, you will get a larger damping of ET due to evaporative cooling. The functional from of $\Delta$ will be the same across PFT, but the temperature range may vary slightly. In contrast, aerodynamic conductance will vary strongly with PFT due to the importance of surface roughness for aerodynamic conductance. So most of the differences in scaling between PFT should be in the aerodynamic conductance term.

@@

\subsection{Bulk statistics of ET response to VPD}
\label{stats_sec}
FORULATE THAT DIFFERNTLY< SAY WE NOW TURN TO OBSERVATIONS> YOU HVAE ALREADY DISCSSED SIGAMA AND IRTS DEFINTION. MENTION IT WHILE HYOU DISCUSS THE RESULS BUT DO NOT MAKE IT THE MAIN FOCUS. I TRIED DOING IT
In this section we consider direct observations of ET response with eddy-covariance data, while including uncertainty, with the $\sigma$ term (Section \ref{methods}). These observational results of ET response largely confirm our theoretical analysis, presented in Sections \ref{sign_term}-\ref{scale_term} (Table \ref{stats}). For all PFTs, mean ET response to increasing VPD is negative. However, ET response evaluated at the average of all variables (e.g. $\sigma$, $T$, $c_s$, $VPD$) is only negative for CSH and GRA. This difference in mean ET response as compared to the ET response at mean environmental conditions is due to the non-linear nature of the response, in which negative responses are generally much larger magnitude than positive responses (Figure \ref{sign_plot}). SAY IT IN WORDS AS WELL AND WHTEHR IT FITS PREVIOUS REULTS

Regarding the frequency of negative and positive ET response, observations of ENF, GRA and CSH are more frequently characterized by a negative ET response to increasing VPD, and are thus more frequently water conservative TOO AVAGUE ESENTENCE REFORMUALTE. DBF and CRO observations have greater frequency of positive ET response to increasing VPD. i.e. water intensive strategies.

In general the bulk statistics match our theoretical expectations well, with the caveat that inclusion of uncertainty shifts grass towards a slightly more negative response than the theory and Figure \ref{sign_plot} would suggest. The bulk statistics motivate a more thorough examination of the structure of uncertainty and more sophisticated validation of our theory's performance against observations.

@@

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
MERGE WITH PREVIOUS PARAGRAPH
We now turn to a direct evaluation of the functional forms presented in Section \ref{sign_term} against observations taken from a wide range of eddy-covariance datasets (Section \ref{methods}). Figure \ref{agu_fig} shows the distribution of of the sign term (when uncertainty is included), as compared to what the theory would predict. Our goal was to capture the leading order behavior of the ET dependence on VPD. Given the assumptions we have made, and the uncertainties of flux tower observations themselves, we expect a relatively large amount of noise when reproducing the derivatives of ET. However, the data largely reproduces our theoretical analysis.

@@

\subsubsection{Testing VPD$_{crit}$}
As stated in the above discussion of ENF, we are mostly interested in the critical threshold $VPD_{crit}$ for which the sign of the ET response to VPD ($\frac{\partial \; ET}{\partial \; VPD}$) is determined. When VPD is less than VPD$_{crit}$, ET response to VPD is negative ($\frac{\partial \; ET}{\partial \; VPD}  < 0$), i.e. plant response dominates, and when VPD is greater than VPD$_{crit}$, ET response to VPD is positive  ($\frac{\partial \; ET}{\partial \; VPD} > 0$). i.e. atmospheric demand dominates. Theoretically, this $VPD_{crit}$ is mostly a function of PFT and not of climate. MAYBE MENTION THAT TEHR SHOULD BE STRONG SPECIES SELCETION BECAUSE OF THIS THRESHOLD

@@

While the stomatal conductance (MAKE SURE YOU SUE CONDUCTANE THROUGHOUT< YOU WERE USING RESITSANE SOMETIMES) s$g_a$ has a large impact on the magnitude of the scaling, temperature, through Clausius-Clapeyron ($\Delta$), does not appear to have as much of an impact in the complete estimation of ET response ($\frac{\partial \; ET}{\partial \; VPD}$). Therefore, it appears that the absolute magnitude of ($\frac{\partial \; ET}{\partial \; VPD}$ can be mostly attributed to the sign term, and its direct scaling by aerodynamic conductance. Intuitively this is reasonable; aerodynamic conductance will control how dominant balances at the land surface are communicated to the boundary layer. While $\Delta$ controls the efficiency of latent heat flux in removing heat from the surface, it appears this is a second order term, relative to $g_a$, for scaling ET response.

@@

\section{Limitations of theory - very dry soil moisture conditions}
In formulating our theory with Penman Monteith, we implicitly did not account for very dry soil moisture conditions. For the majority of environmental conditions  observed at the eddy covariance sites used here, soil conditions were not extremely dry so that we could assume a constant $uWUE$ and $g_1$. We posit that ecosystems will generally optimize to host plants living in conditions which they evolved for. However, in extreme conditions and drought scenarios soil water content (SWC) could become the limiting factor for ET response to VPD, which our theory does not account for. IN addition low soil moisture conditions themselves increase PET and VPD through land-atmosphere feedback (Berg NAT CL CC, BOUCHET MRTON< BRUTSAERT< LINTENER AND GENTINE Budkyko and complementary relatnship). We first note that in our datasets those conditions very very infrequent (talk about pdf figures). Within our framework, any systematic bias due to the  failure to account for SWC's effects in very dry conditions should manifest itself in a functional relationship between $\sigma$, our uncertainty measure, and SWC, which is observed at the FLUXNET sites. Examining the relationship between $\sigma$ and SWC will test to what extent our theory breaks down in very dry soil moisture conditions.

If we view $\sigma \cdot uWUE$ as, in addition to a measure of uncertainty, a short time-scale observation of uWUE, we would expect $\sigma \cdot uWUE$ to decrease at low SWC (ZHOU ET AL). If $\sigma \cdot uWUE$ is a very strong function of SWC, then our theory should be conditioned more strongly on well watered soil conditions. If $\sigma \cdot uWUE$ is weakly a function of SWC, then our theory is more universal and independent of soil moisture conditions.

Indeed, for all PFTs, there is some slight dependence of $\sigma \cdot uWUE$ on SWC, especially at low SWC (Figure \ref{swc}). The portion of observations for which our theory is biased by SWC-limitations varies by PFT, due to the nonlinear threshold at which soil moisture availability limits plant function. For CRO and DBF, soil water content only matters for about the lowest 5\% of observations (each box is 5\% of observations in Figure \ref{swc}). And, CRO, relative to DBF, has a very weak dependence on soil moisture, which is reflective of the high likelihood that CRO sites are optimally irrigated and not water stressed. For ENF, systematic SWC-induced biases in $\sigma \cdot uWUE$ emerge in about the lowest 30\% of SWC conditions. In crontast to ENF, CRO, and DBF, for GRA the relationship between $\sigma \cdot uWUE$ and SWC is more linear and affects a greater portion of observations; about 60\% of observations. Clearly, for GRA soil water frequently impacts plant function and alters ET response, and our theory is limited. It is therefore not suprising that for GRA our theory tested the worst (ARE YOU SURE ABOUT THIS WORD) againt the data (Section \ref{testing}); for all other PFTs occurence of soil mosture impacts was rare enough to not manifest itself in bulk statistics and plots\footnote{note that interpreting the soil moisture dependence for CSH is difficult as their are only two sites for that PFT. So systematic soil moisture relationshps are not apparent in Figure \ref{swc}, but instead blurred by the superposition of the two sites. We argue taht the soil moisture dependence is still there, but we don't have enough statistcs to comment on it}.

MERGE WITH THE SENTENCE I PLACED ABOVE WITH LAND-ATMOPHERE COUPLPINGThe observed dependence of $\sigma \cdot uWUE$ on SWC for GRA would explain the deficiencies of our theory compared to the observations in Section \ref{testing}, specifically the trend back towards negative soil moisture at high VPD. Feedbacks between the land surface and the atmosphere, which our not accounted for due to our focus on the one-way response of the land surface to atmospheric conditions, would cause high VPD to be correlated with low SWC \citep[][Gentine et al. GRL 2016]{Berg_2016}\explain{Pierre, which publication were you thinking, your Evaporation from Weather Data, YES?}. So, at high VPD observations of low SWC are more likely, and this low SWC causes a lower uWUE (ZHOU ET AL). The lower uWUE at low SWC/high VPD then leads to the observed downturn towards decreasing ET with increasing VPD, and the deviation between our theory (based on a constant uWUE assumption) and observations.

@@

We derived a new form of Penman Monteith using the concept of semi-empirical optimal stomatal regulation (CITE LIN AND MEDLYN) and near constant $uWUE$ (\cite{Zhou_2015}) to remove the stomatal conductance implicit dependence on GPP and ET. With our new form of Penman Monteith we developed a theory for when an ecosystem will tend to reduce or increase ET with increasing VPD, which we evaluated against a range of eddy-covariance data spanning different climates and plant functional types. The goal was to capture the leading order behavior of the system to gain some intrinsic knowledge for its behavior. This intuition can be used to disentangle land atmosphere feedbacks in more complicated scenarios, and will aid interpretation of observations and sophisticated models.

Our theory states that each plant type has a critical VPD below which ET will decrease (plant response dominating) and above which ET will increase (atmospheric demand dominates). We tested our theory using data from FLUXNET 2015, and found that for DBF and CSH the theory succeeds in capturing the leading order behavior. For ENF, the theory succeeds in capturing the leading order behavior for the sign of $\frac{\partial \; ET}{\partial \; VPD}$, but not for magnitude. In contrast, the model does not capture the behavior for eddy-covariance CRO and GRA, which could have been expected, as our theory does not account for large sources of heterogeneity in those plant types, including varying surface roughness, c3 vs c4 photosynthesis, and potential differences in irrigation across sites. Also, our theory does not account for strong soil drying, which non-negligibly alters ET response for GRA. Given these deficiencies for CRO and GRA, we will withhold conclusions on their response to increasing VPD.

For the forest sites (DBF, ENF) the observed environmental VPD approximately straddles the critical VPD, with about half the observations below the critical VPD and half above. However for CSH the environmental VPD never exceeds the critical VPD, and even when uncertainty is included the sign of the derivative is almost always negative (93\% of observations).
MAYBE HAVE A SIMPLE CONCLUDING STATEMETN REGARDING THE FACT THAT MOST PLANTS ARE CONSERVATIVE EXCEPT FOR CRO NA D THAT STOMATAL REGULATION STONGLY REGULATES ET. BASICALLY ANSWER THE TITEL QUESTION.

While our theory predicts common occurrence of positive $\frac{\partial \; ET}{\partial \; VPD}$ (atmospheric demand dominates), the plant regulation is still important even when $\frac{\partial \; ET}{\partial \; VPD} > 0$. The magnitude of the ``sign term'' of our theory is always far below what it would be if we only considered the atmospheric demand term (PET). This means that any drought indices or analysis using PET will not capture how the vegetated surface respond and react to increases in atmospheric demand. However, the new uWUE-version of Penman Monteith we derived (Equation \ref{et}) could be used as a complement to PET in drought indices. One could calculate and compare indices  when plant response is ignored (PET) or included (Equation \ref{et}).

>>>>

$\Delta$'s presence in the scaling term also matches physical intuition. Evaporative cooling will dampen the ability of the atmosphere to take more moisture (UNCLEAR SENTENCE REFORMULATE **), because $e_{s}$ decreases with decreasing temperature. The decrease in $e_{s}$ is proportional to $\Delta$ ($\delta e_{s} = \Delta \delta T$). So as $\Delta$ increases, you will get a larger damping of ET due to evaporative cooling. The functional from of $\Delta$ will be the same across PFT, but the temperature range may vary slightly. In contrast, aerodynamic conductance will vary strongly with PFT due to the importance of surface roughness for aerodynamic conductance. So most of the differences in scaling between PFT should be in the aerodynamic conductance term.

@@

\subsection{Bulk statistics of ET response to VPD}
\label{stats_sec}
FORULATE THAT DIFFERNTLY< SAY WE NOW TURN TO OBSERVATIONS> YOU HVAE ALREADY DISCSSED SIGAMA AND IRTS DEFINTION. MENTION IT WHILE HYOU DISCUSS THE RESULS BUT DO NOT MAKE IT THE MAIN FOCUS. I TRIED DOING IT

In this section we consider direct observations of ET response with eddy-covariance data, while including uncertainty with the $\sigma$ term (Section \ref{methods}). These observational results of ET response largely confirm our theoretical analysis, presented in Sections \ref{sign_term}-\ref{scale_term} (Table \ref{stats}). For all PFTs, mean ET response to increasing VPD is negative. However, ET response evaluated at the average of all variables (e.g. $\sigma$, $T$, $c_s$, $VPD$) is only negative for CSH and GRA. This difference in mean ET response as compared to the ET response at mean environmental conditions is due to the non-linear nature of the response, in which negative responses are generally much larger magnitude than positive responses (Figure \ref{sign_plot}). Therefore, both the mean ET response as well as the ET response at mean environmental conditions matches our expectations from the theory (Section \ref{sign_term}), with the exception that CRO observations are shifted more towards a negative ET response than we expect.

Regarding the frequency of negative and positive ET response, 53\% of ENF observations, 65\% of GRA observations, and 93\% of CSH observations are characterized by a negative ET response to increasing VPD. The response for these PFTs is thus more frequently physiologicaly controlled, and these PFTs will tend to conserve soil water for futre use. However DBF and CRO observations have greater frequency of positive ET response to increasing VPD (46\% and 49\% of observations, respectively), and are more intensive in their use of water.

In general the bulk statistics match our theoretical expectations well, with the caveat that inclusion of uncertainty shifts crops towards a slightly more negative response than the theory and Figure \ref{sign_plot} would suggest. The bulk statistics motivate a more thorough examination of the structure of uncertainty and more sophisticated validation of our theory's performance against observations.

@@

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
While Section \ref{stats_sec} presented simple statistics of observed response, we now compare more sophisticated distributions of the observed response to our theory presented in Section \ref{sign_term}. Figure \ref{agu_fig} shows observed the distribution of of the sign term, as compared to what the theory would predict. Our goal was to capture the leading order behavior of the ET dependence on VPD. Given the assumptions we have made, and the uncertainties of flux tower observations themselves, we expect a relatively large amount of noise when reproducing the derivatives of ET. However, the data largely reproduces our theoretical analysis.

@@

\subsubsection{Testing VPD$_{crit}$}
As stated in the above discussion of ENF, we are mostly interested in the critical threshold $VPD_{crit}$ for which the sign of the ET response to VPD ($\frac{\partial \; ET}{\partial \; VPD}$) is determined. When VPD is less than VPD$_{crit}$, ET response to VPD is negative ($\frac{\partial \; ET}{\partial \; VPD}  < 0$), i.e. plant response dominates, and when VPD is greater than VPD$_{crit}$, ET response to VPD is positive  ($\frac{\partial \; ET}{\partial \; VPD} > 0$). i.e. atmospheric demand dominates. Theoretically, this $VPD_{crit}$ is mostly a function of PFT and not of climate. This $VPD_{crit}$, and how it compares with environmental VPD and water supply, will exert a strong control on species selection in competitive environments.

@@

While the stomatal conductance $g_a$ has a large impact on the magnitude of the scaling, temperature, through Clausius-Clapeyron ($\Delta$), does not appear to have as much of an impact in the complete estimation of ET response ($\frac{\partial \; ET}{\partial \; VPD}$). Therefore, it appears that the absolute magnitude of ($\frac{\partial \; ET}{\partial \; VPD}$ can be mostly attributed to the sign term, and its direct scaling by aerodynamic conductance. Intuitively this is reasonable; aerodynamic conductance will control how dominant balances at the land surface are communicated to the boundary layer. While $\Delta$ controls the efficiency of latent heat flux in removing heat from the surface, it appears this is a second order term, relative to $g_a$, for scaling ET response.

@@

\section{Limitations of theory - very dry soil moisture conditions}
In formulating our theory with Penman Monteith, we implicitly did not account for very dry soil moisture conditions. For the majority of environmental conditions  observed at the eddy covariance sites used here, soil conditions were not extremely dry so that we could assume a constant $uWUE$ and $g_1$. We posit that ecosystems will generally optimize to host plants living in conditions which they evolved for. However, in extreme conditions and drought scenarios soil water content (SWC) could become the limiting factor for ET response to VPD, which our theory does not account for. In addition, low soil moisture conditions themselves increase VPD through land-atmosphere feedback \citep[][]{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006, Salvucci_2013, Gentine_2016, Berg_2016}.

We first note that in our datasets low SWC conditions are very infrequent (MAKE SEPERATE PDF OF SWC?). Within our framework, any systematic bias due to the  failure to account for SWC's effects in very dry conditions should manifest itself in a functional relationship between $\sigma$, our uncertainty measure, and SWC, which is observed at the FLUXNET sites. Examining the relationship between $\sigma$ and SWC will test to what extent our theory breaks down in very dry soil moisture conditions.

If we view $\sigma \cdot uWUE$ as, in addition to a measure of uncertainty, a short time-scale observation of uWUE, we would expect $\sigma \cdot uWUE$ to decrease at low SWC (CITE ZHOU ** which ZHOU shows low uWUE with low SWC). If $\sigma \cdot uWUE$ is a very strong function of SWC, then our theory should be conditioned more strongly on well watered soil conditions. If $\sigma \cdot uWUE$ is weakly a function of SWC, then our theory is more universal and independent of soil moisture conditions.

Indeed, for all PFTs, there is some slight dependence of $\sigma \cdot uWUE$ on SWC, especially at low SWC (Figure \ref{swc}). The portion of observations for which our theory is biased by SWC-limitations varies by PFT, due to the nonlinear threshold at which soil moisture availability limits plant function. For CRO and DBF, soil water content only matters for about the lowest 5\% of observations (each box is 5\% of observations in Figure \ref{swc}). And, CRO, relative to DBF, has a very weak dependence on soil moisture, which is reflective of the high likelihood that CRO sites are optimally irrigated and not water stressed. For ENF, systematic SWC-induced biases in $\sigma \cdot uWUE$ emerge in about the lowest 30\% of SWC conditions. In crontast to ENF, CRO, and DBF, for GRA the relationship between $\sigma \cdot uWUE$ and SWC is more linear and affects a greater portion of observations; about 60\% of observations. Clearly, for GRA soil water frequently impacts plant function and alters ET response, and our theory is limited. It is therefore not suprising that our theory tested poorly againt the data for GRA, relative to ENF, DBF, and GRA (Section \ref{testing}). For all other PFTs occurence of soil mosture impacts was rare enough to not manifest itself in bulk statistics and plots\footnote{note that interpreting the soil moisture dependence for CSH is difficult as their are only two sites for that PFT. So systematic soil moisture relationshps are not apparent in Figure \ref{swc}, but instead blurred by the superposition of the two sites. We argue taht the soil moisture dependence is still there, but we don't have enough statistcs to comment on it}.

The observed dependence of $\sigma \cdot uWUE$ on SWC for GRA would explain the deficiencies of our theory compared to the observations in Section \ref{testing}, specifically the trend back towards negative soil moisture at high VPD. Aforementioned feedbacks between the land surface and the atmosphere, which are not accounted for due to our focus on the one-way response of the land surface to atmospheric conditions, would cause high VPD to be correlated with low SWC \citep[][]{Getine_2016, Berg_2016}. So, at high VPD observations of low SWC are more likely, and this low SWC causes a lower uWUE (ZHOU ET AL **). The lower uWUE at low SWC/high VPD then leads to the observed downturn towards decreasing ET with increasing VPD, and the deviation between our theory (based on a constant uWUE assumption) and observations.

@@

We derived a new form of Penman Monteith using the concept of semi-empirical optimal stomatal regulation \citep{Lin_2015, MEDLYN_2011} and near constant $uWUE$ \citep{Zhou_2015} to remove the stomatal conductance implicit dependence on GPP and ET. With our new form of Penman Monteith we developed a theory for when an ecosystem will tend to reduce or increase ET with increasing VPD, which we evaluated against a range of eddy-covariance data spanning different climates and plant functional types. The goal was to capture the leading order behavior of the system to gain some intrinsic knowledge for its behavior. This intuition can be used to disentangle land atmosphere feedbacks in more complicated scenarios, and will aid interpretation of observations and sophisticated models.

Our theory states that each plant type has a critical VPD below which ET will decrease (plant response dominating) and above which ET will increase (atmospheric demand dominates). We tested our theory using data from FLUXNET 2015, and found that for DBF and CSH the theory succeeds in capturing the leading order behavior. For ENF, the theory succeeds in capturing the leading order behavior for the sign of $\frac{\partial \; ET}{\partial \; VPD}$, but not for magnitude. In contrast, the model does not capture the behavior for eddy-covariance observations of CRO and GRA, which could have been expected, as our theory does not account for large sources of heterogeneity in those plant types, including varying surface roughness, c3 vs c4 photosynthesis, and potential differences in irrigation across sites. Also, our theory does not account for strong soil drying, which non-negligibly alters ET response for GRA. Given these deficiencies for CRO and GRA, we will withhold conclusions on their response to increasing VPD.

For the forest sites (DBF, ENF) the observed environmental VPD approximately straddles the critical VPD, with about half the observations below the critical VPD and half above. However for CSH the environmental VPD never exceeds the critical VPD, and even when uncertainty is included the sign of the derivative is almost always negative (93\% of observations).

Generally speaking, our theory suggests that for a majority of enviornmental VPD conditions, plants will tend to conserve water and reduce ET with increasing VPD. Stomatal regulation and plant physiological response strongly regulates ET, but his regulation does vary by PFT. CROs are the least water conservative, while DBF, ENF, and GRA are progressively more water conservative, with CSH begin the most water conservative (most likely to reudce ET in response to increasing VPD). However, these results are subject to the caveat that our theory's performance is most uncertain for GRA and CRO.

While our theory predicts occurrence of increasing ET with increasing VPD (atmospheric demand dominates), plant regulation is still important even when $\frac{\partial \; ET}{\partial \; VPD} > 0$. The magnitude of the ``sign term'' of our theory is always far below what it would be if we only considered the atmospheric demand term (PET). This means that any drought indices or analysis using PET will not capture how the vegetated surface respond and react to increases in atmospheric demand. However, the new uWUE-version of Penman Monteith we derived (Equation \ref{et}) could be used as a complement to PET in drought indices. One could calculate and compare indices  when plant response is ignored (PET) or included (Equation \ref{et}).
